---
title: "Analysis of Property Data Set"
author: "Justin Gatonby"
date: "`r Sys.Date()`"
output:
  word_document:
    reference_docx: Data_Analysis.docx
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load_packages, include = FALSE}
# Tidyverse includes dplyr and ggplot2 packages 
library(tidyverse)
# Additional packages used only in certain functions 
library(ggtext)
library(ggmap)
library(patchwork)
```


```{r get_data, include = FALSE}
data <- read.csv("property_data(2).csv", header = TRUE) 
```

```{r clean_data, include = FALSE}
# If the observation is classified as Vacant Land, 
# bedrooms, bathrooms, and parkings spaces will be zero 
data <- data %>% 
  mutate(
    Bedrooms = ifelse(Type == "land", 0, Bedrooms), 
    Bathrooms = ifelse(Type == "land", 0, Bathrooms), 
    Parking.Spaces = ifelse(Type == "land", 0, Parking.Spaces) 
  )
# If bedrooms or bathrooms are missing, remove observation from data set
data <- data %>%
  filter(!is.na(Bathrooms) & !is.na(Bedrooms))
# Assume if parking spaces is missing, there os no parking available 
data <- data %>% 
  mutate(Parking.Spaces = ifelse(is.na(Parking.Spaces), 0, Parking.Spaces))
# Manually add coordinates to those not found in coords.py 
# Otherwise leave existing coordinates as given 
data <- data %>%
  mutate(
    Latitude = ifelse(Location == "Costa Da Gama", -34.09616,
               ifelse(Location == "Wynberg Upper", -34.00685,
               ifelse(Location == "Browns Farm", -34.03064,
               ifelse(Location == "Claremont Upper", -33.98342, 
               ifelse(Location == "Clarkes Estate", -33.93851,
               ifelse(Location == "Greenville Garden City", -33.78286,
               ifelse(Location == "Kenilworth Upper", -33.99487, 
               ifelse(Location == "Montana", -33.95787,
               ifelse(Location == "Ottery East", -34.01786, Latitude))))))))),  
    Longitude = ifelse(Location == "Costa Da Gama", 18.48278,
                ifelse(Location == "Wynberg Upper", 18.46940,
                ifelse(Location == "Browns Farm", 18.51234,
                ifelse(Location == "Claremont Upper", 18.46048, 
                ifelse(Location == "Clarkes Estate", 18.58340, 
                ifelse(Location == "Greenville Garden City", 18.72129, 
                ifelse(Location == "Kenilworth Upper", 18.47379, 
                ifelse(Location == "Montana", 18.58109,
                ifelse(Location == "Ottery East", 18.51088, Longitude))))))))))
# Any observations still missing coordinates are removed 
data <- data %>%
  filter(!is.na(Latitude) & !is.na(Longitude))
# Remove observations where both Floor Size and Erf Size are missing
data <- data %>%
  filter(!is.na(Floor.Size) | !is.na(Erf.Size))
# A single value needs to be removed, must have been an error in data collection 
data <- data %>%
  filter(Type != "l>\n")
data <- data %>%
  filter(Location != "Nooitgedacht")
```


```{r type, echo = FALSE, fig.width=8, fig.height=6, dpi=300}
# Add features related to Price and order for visualization
data <- data %>%
  group_by(Type) %>%
  mutate(
    Type = str_to_title(Type), 
    Median_price = median(Price), 
    Type_with_count = paste0(Type, "(", n(), ")")
  ) %>%
  ungroup() %>%
  mutate(Type_with_count = fct_reorder(Type_with_count, Median_price))
# Median price of the data set 
median_price <- median(data$Price)
# Create box-plot 
ggplot(data, aes(x = Type_with_count, y = Price, fill = Type)) + 
  geom_boxplot(color = "black") + 
  geom_hline(yintercept = median_price) +
  scale_fill_brewer(palette = "Set1") + 
  labs(title = "Property Prices by Type", 
       x = "Property Type (with Counts)",
       y = "Price (ZAR)"
       ) + 
  stat_summary(fun = median, geom = "text", aes(label = round(after_stat(y), 0)), 
             vjust = -0.5, color = "black", size = 3) + 
  theme_light() + 
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_markdown(
      hjust = 0.5,
      face = "bold",
      size = 16,
      margin = margin(b = 10)
    ),
  )
# Remove the variable "Median_price" from the data set
data <- data %>% select(-Median_price, -Type_with_count)
```

# Analysis of Property Prices by Type 

**Price Trends**

* Indicates properties with greater amenities, development, and complexity command higher prices.
* Suggests apartments and townhouses are typically more expensive than vacant land and houses. 

\newpage

**Distribution**

* Vacant land has a symmetrical distribution, with a large range relative to its interquartile range, suggesting that there are outliers on both ends (overvalued and undervalued).
* Houses have the largest interquartile range, suggesting that other variables will be needed to determine their price. 
* Apartments have a long tail extending towards lower prices, displaying an opportunity to identify undervalued properties.
* Townhouses are negatively skewed, with fewer townhouses priced significantly lower than the median price and fewer opportunities to find undervalued properties. 

**Composition**

* The bulk of the data set is composed of apartments (66,4%), followed by houses (28,2%).
* Land (2,5%) and townhouses (2,8%) make up a far smaller portion of the dataset. 

\newpage

```{r location, echo = FALSE, fig.width=8, fig.height=6, dpi=300}
# 6 locations to help orient the viewer 
key_locations <- data %>%
  group_by(Location) %>%
  filter(n() >= 3) %>%
  ungroup() %>%
  distinct(Location, Latitude, Longitude) %>%
  slice_head(n = 6) # Select first 6 meeting above criteria
# Add Greenville Garden City 
greenville <- data %>%
  filter(Location == "Greenville Garden City") %>%
  distinct(Location, Latitude, Longitude) %>%
  mutate(Location = "Greenville\nGarden City") # Add a newline for better layout
# Add Sea Point
sea_point <- data %>%
  filter(Location == "Sea Point") %>%
  distinct(Location, Latitude, Longitude)
# Add Zonnebloem
zonnebloem <- data %>%
  filter(Location == "Zonnebloem") %>%
  distinct(Location, Latitude, Longitude)
# Combine key locations
key_locations <- bind_rows(key_locations, greenville, sea_point, zonnebloem)
# Create scatter plot with labels
ggplot(data, aes(x = Latitude, y = Longitude)) +
  geom_jitter(aes(color = Price), size = 2, alpha = 0.7, width = 0.01, height = 0.01) +
  geom_text(
    data = key_locations,
    aes(x = Latitude, y = Longitude, label = Location),
    size = 3, fontface = "bold", color = "black", nudge_y = 0.01
  ) +
  scale_color_viridis_c(option = "C", direction = 1, name = "Price (ZAR)") +
  labs(
    title = "Property Prices by Location",
    subtitle = "Plotted by Coordinates",
    x = "Latitude",
    y = "Longitude"
  ) +
  theme_light() +
  theme(
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
    plot.subtitle = element_text(size = 12, hjust = 0.5),
    legend.position = "right"
  )
```

# Analysis of Property Prices by Location

**Price Trends**

* Location is a strong predictor of property price. 
* Areas around the coast tend to be more expensive, while those further away are on the lower end of the price range. 

**Distribution and Composition**

* Most properties in this data set are clustered around the coast. 
* As longitude increases, measuring similarity becomes more challenging, as property dispersion and price variability increases. 

\newpage

```{r bed/bathrooms, echo = FALSE, fig.width=8, fig.height=6, dpi=300}
# Removing outliers found from first plot
new_data <- data %>% 
  filter(Bedrooms < 8 & Bathrooms < 5)
# Create scatter plot with jittered points 
ggplot(new_data, aes(x = Bedrooms, y = Bathrooms)) +
  geom_jitter(aes(color = Price), width = 0.1, height = 0.1, alpha = 0.7) +
  scale_color_viridis_c(option = "C", direction = 1, name = "Price (ZAR)") + 
  labs(
    title = "Property Prices by Bedrooms and Bathrooms", 
    x = "Bedrooms", 
    y = "Bathrooms"
  ) + 
  theme_light() + 
  theme(
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5)
  )

```

# Analysis of Property Prices by Bedrooms and Bathrooms

**Price Trends**

* Properties with more bedrooms and bathrooms tend to demand higher prices.
* However, more bedrooms will only correspond with higher prices, if the number of bathrooms increases as well. 


**Distribution**

* Most points are clustered around combinations of 1-3 bedrooms and 1-2 bathrooms.

**Outliers**

* Extreme observations have been removed from the plot, that is, properties with more than 7 bedrooms or more than 4 bathrooms.
* Zero bedrooms and bathrooms is not an anomaly in the data, but rather an indicator of vacant land. 

\newpage 

```{r size, echo = FALSE, fig.width=8, fig.height=6, dpi=300}
# Clean data to plot erf size 
erf_data <- data %>%
  filter(!is.na(Erf.Size))
# Clean data to plot floor size
floor_data <- data %>%
  filter(!is.na(Floor.Size))
# Scatter plot for Price vs Erf.Size
erf_size_plot <- ggplot(erf_data, aes(x = Erf.Size, y = Price)) +
  geom_point(alpha = 0.7, color = "blue") +
  labs(
    title = "Price vs Erf Size",
    x = expression("Erf Size (m"^2*")"),
    y = "Price (ZAR)"
  ) +
  theme_light()
# Scatter plot for Price vs Floor.Size
floor_size_plot <- ggplot(floor_data, aes(x = Floor.Size, y = Price)) +
  geom_point(alpha = 0.7, color = "green") +
  labs(
    title = "Price vs Floor Size",
    x = expression("Floor Size (m"^2*")"),
    y = "Price (ZAR)"
  ) +
  theme_light()

# Combine the plots using patchwork
combined_plot <- erf_size_plot + floor_size_plot + 
  plot_annotation(
    title = "Property Prices by Erf Size and Floor Size",
    theme = theme(plot.title = element_text(size = 16, face = "bold", hjust = 0.5))
  )

# Print the combined plot
combined_plot

# print(cor(floor_data$Price, floor_data$Floor.Size))
# print(cor(erf_data$Price, erf_data$Erf.Size))
```

# Analysis of Property Prices by erf and floor size 

**Correlation**

* Both erf size and floor size show a positive correlation with price, with larger properties commanding higher prices. 
* The strength of the relationship between erf size and price (r=0.26) and floor size and price (r=0.24) is similar.

**Outliers**

* There are many outliers in both plots at both ends.
* Many small properties are priced at the top of the price range, while some of the largest properties are at the lower end.

**Reliability**

* Although there is some correlation between each variable and price, the weakness of the correlation and the number of outliers, make size on its own a weak predictor of property price. 


# Summary 

* The coordinates of a property appear to be the best predictor of its price. 
* Therefore, a k-nearest neighbor model should produce accurate predictions of property price. 
* The model may use other informative variables such as number of bedrooms, number of bathrooms, erf size, floor size and number of parking spaces as well.
* Categorical variables may be excluded to avoid the "Curse of Dimensionality" and challenges converting them to numeric features.
* This should not affect model performance as the location is described by coordinates, while the number of bedrooms and bathrooms are indicative of property type. 
